### 2021-07-06 rad-icfp-2021/complexity.tex
changing outline
simple writing phase
bit more related work
minor

\subsection{Cost model}

@@

\MH{need cost array operations}
\MH{need to extend the cost function in a compositional way}

@@

>>>>

\subsection{Cost model}
\label{sub:costModel}

@@

TODO: extend cost array
TODO: extend cost compositional way

@@

The cheap gradient principle (see e.g. \cite{griewank2008evaluating}) for reverse-mode
asserts that evaluating the gradient of a function $e:\RR^n\to\RR$ 
should be the same order of cost as evaluating $e$. 
More precisely, there should be a constant $K$ such that for each program !$\Gamma \vdash$ e: $\reals$! in the context $\Gamma=\{x_1:\reals,\ldots,x_n:\reals\}$,
 !$\cost$($\grad$e) $\leq$ $K$*$\cost$(e)!.

As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle. 
However, this is not quite the case with $\directD{\rho}{\Gamma}{Y}$ 
before the optimization (inlining and forward substitution) step from Section~\ref{sub:Partial evaluation and optimization}.

\begin{proposition}
    TODO: cost $\directD{\rho}{\Gamma}{Y}$. 
\end{proposition}

As we can see, the only problem is that too many variables are being moved, 
due to the continuation carrying and copying $n$ variables at each step.
This is exactly the overhead that our inlining and forward substitution removes.

\begin{proposition}
    TODO: complexity gain using our optim.
\end{proposition}

Combining the two previous points, we conclude that our reverse mode transformation satisfies the cheap gradient principle.

\begin{theorem}
    TODO: The reverse mode transformation satisfies the cheap gradient principle.
\end{theorem}

\subsection{Optimizations} % (fold)
\label{sub:Optimizations}

In Figure~\ref{fig:optim} we present a of optimizations. 
As our language is purely functional and effect free, 
we can use these optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let bindings 
and resembles SSA-code.

\input{figures/optims}

### 2021-07-06 rad-icfp-2021/correctness.tex
small fixes
minor

\begin{definition}[$\CartSp$]
Let $\CartSp$ be the Cartesian category whose objects are Euclidean spaces
and whose morphisms are smooth functions between Euclidean spaces.
\MH{it's not a multicategory yet}

@@

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$
and $reduce_{x,y.e1;e2}$.
\MH{actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.}
\end{definition}

@@

\begin{definition}[$\ConcatT$]
Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
with primitives given by 
$op1: \reals \to \reals$, 
$op2: \reals,\reals \to \reals$, 
$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$, 
$reduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
$J^Top1: \reals \to \reals$, 
$J^Top2: \reals \to \reals, \reals$, 
$J^Tmap2_{x,y.e}: \Array{\reals}{n} \to \Array{\reals}{n}, \Array{\reals}{n}$, 
$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
\MH{not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?}
\end{definition}

@@

\MH{I think need to explain here the (co)monoid structure induced by $D$}

\begin{tabular}{l c l}
    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& TODO \\
    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=&  \\
    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& \\
    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& \\
    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& \\
    $\seml$!J!$^T$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
    $\seml$!<e1, e2>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
    $\seml$!e1!$\icomp$!e2!$\semr$ &=&  \\
 \end{tabular}
\MH{Do I need any kind of proposition saying that this makes sense?
Might need to define on types as well to be clearer: It's an identity on types functor...}

@@

\MH{I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.}

>>>>

\begin{definition}[$\CartSp$]
Let $\CartSp$ be the Cartesian category whose objects are Euclidean spaces
and whose morphisms are smooth functions between Euclidean spaces.
TODO:it's not a multicategory yet

@@

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$ $map2_{x,y.e}$
and $reduce_{x,y.e1;e2}$.
TODO: actually I just need projection and a representing object for pairs, it might be weaker than Cartesian? edit: I think I need Cartesian.
\end{definition}

@@

\begin{definition}[$\ConcatT$]
Let $\ConcatT$ be the syntactic Cartesian concategory whose types are those of Target and 
with primitives given by 
$op1: \reals \to \reals$, 
$op2: \reals,\reals \to \reals$, 
$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$, 
$reduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
$J^Top1: \reals \to \reals$, 
$J^Top2: \reals \to \reals, \reals$, 
$J^Tmap2_{x,y.e}: \Array{\reals}{n} \to \Array{\reals}{n}, \Array{\reals}{n}$, 
$J^Treduce_{x,y.e1;e2}: \Array{\reals}{n} \to \Array{\reals}{n}$, 
TODO: not too sure, should I say what a Cartesian concat is? do I need $J^Tproj$?
\end{definition}

@@

TODO: I think need to explain here the (co)monoid structure induced by $D$

\begin{tabular}{l c l}
    $\seml$!J!$^T$!var!$_{T;i} \semr$ &=& TODO \\
    $\seml$!J!$^T$!op!$_{T;n} \semr$ &=&  \\
    $\seml$!J!$^T$!pair!$_{T;A\times B} \semr$ &=& \\
    $\seml$!J!$^T$!proj!$_{T1;T2;T3} \semr$ &=& \\
    $\seml$!J!$^T$!map2!$_{T;x,y.e}\semr$  &=& \\
    $\seml$!J!$^T$!reduce!$_{T;x,y.e;e}\semr$  &=& \\
    $\seml$!<e1, e2>!$\semr$  &=& !<!$\seml$!e1!$\semr$, $\seml$!e2!$\semr$!>! \\
    $\seml$!e1!$\icomp$!e2!$\semr$ &=&  \\
 \end{tabular}
TODO: Do I need any kind of proposition saying that this makes sense?
Might need to define on types as well to be clearer: It's an identity on types functor...

@@

TODO: I don't want to talk too much about the cotangent bundle functor, but to phrase correctness properties clearly I might have to.

### 2021-07-06 rad-icfp-2021/intro.tex
minor

\MH{TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.}

@@

\MH{TODO: same as above, maybe rewrite for more theory audience.}

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalization}) \MH{should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.}
\end{itemize}

>>>>

TODO: rewrite for more theory people. some ideas: for canonical than Plotkin's thing because normal semantics, complexity proof compared to all other recent papers, 
all papers that have implementations, their interface in the paper is too high level and misses a lot of things (e.g. in \cite{vytiniotis2019differentiable,lantern_icfp,sherman2021}) compared to implementation, hinders progress. 
For instance, in \cite{lantern_icfp} it's a lot of manually optimized and hand-derived primitives to achieve good performance.

@@

TODO: same as above, maybe rewrite for more theory audience.

@@

Our contribution makes these claims precise. In more details, we introduce
\begin{itemize}
    \item an expressive source language with array operations, yet restrictive enough for efficient differentiation (\S\ref{sub:sourcelang})
    \item a method for extending our work and modular approaches to add and prove correct new primitives as needed in practice (\S\ref{sec:generalization})
    \item a new intermediate representation (IR), UNF, essentially turning every operation into a unary one which inputs and outputs a tuple (\S\ref{sec:unf})
    \item a transformation between the IR and the main language (\S\ref{sub:transformations to and from UNF})
    \item a simple reverse-mode transformation on the IR (\S\ref{sub:Simple reverse mode transformation})
    \item the direct reverse-mode transformation obtained from the source to the target language (\S\ref{sub:Macro for pure reverse mode transformation})
    \item a denotational semantics of our languages and transformations using multicategories and concategories (\S\ref{sec:correctness})
    \item correctness of the transformations (\S\ref{sec:correctness})
    \item complexity guarantees (\S\ref{sec:complexity})
    \item how efficiency is ensured via standard optimization techniques (\S\ref{sub:Optimizations})
    \item several additions to the source language (\S\ref{sec:generalization}) TODO:should emphasize that it should be easier for us to add primitives compared to several recent work. give recipe.
\end{itemize}

### 2021-07-06 rad-icfp-2021/notes/main.tex
minor syntax
minor

\begin{figure}[H]
    \setlength{\tabcolsep}{0.3em}
    \centering
    \begin{tabular}{|l c l|l|}
    \hline
    \multicolumn{3}{|c|}{\textbf{Core Grammar}} & \multicolumn{1}{c|}{\textbf{Description}}\\\hline
    !T! & \mbox{::=} & ... & \grammarcomment{Same as source} \\
    \hline
    !e! & \mbox{::=} & !var!$_{\Gamma,i}$ & \grammarcomment{Variable}\\
    & $\mid$ & !op!$_{\Gamma,n}$ & \grammarcomment{Operations, for $0\leq n\leq 2$}\\
    & $\mid$ & !e;e! & \grammarcomment{Sequential composition}\\
    & $\mid$ & !map!$_{\Gamma,x.e}$ $\mid$ !map2!$_{\Gamma,x,y.e}$ & \grammarcomment{Map and map2}\\
    & $\mid$ & !foldl!$_{\Gamma,x,y.e}$ & \grammarcomment{Fold left}\\
    \hline
    \end{tabular}
    \vspace{-0.2cm}
    \caption{Grammar of the source UNF}
    \label{fig:unf_source_grammar}
    \end{figure}

@@

 \MH{I probably need to add permutations $\sigma$ in the language of UNF then.}

@@

\MH{another possible way to look at it is that UNF does some sort of localization, turning every operator to a reversible one, which is simply done by making it also return its context. I wonder if it could simply be seen as the localization of $!$. Assuming $1\times A=A$, we have $\Delta;id\times !=id$, so as $!$ would become invertible, so would be $\Delta$. This is sort of what I need in practice for reverse mode: not seeing any $\Delta$, only unary maps, and an iso can always be seen as such in this context, i.e. by making $\Delta$ an iso, it becomes essentially invisible to reverse mode. One obvious problem is that I don't know if I want all types to be isomorphic to 1.}

>>>>

\begin{figure}[H]
    \setlength{\tabcolsep}{0.3em}
    \centering
    \begin{tabular}{|l c l|l|}
    \hline
    \multicolumn{3}{|c|}{\textbf{Core Grammar}} & \multicolumn{1}{c|}{\textbf{Description}}\\\hline
    !T! & \mbox{::=} & $\ldots$ & \grammarcomment{Same as source} \\
    \hline
    !e! & \mbox{::=} & !var!$_{\Gamma,i}$ & \grammarcomment{Variable}\\
    & $\mid$ & !op!$_{\Gamma,n}$ & \grammarcomment{Operations, for $0\leq n\leq 2$}\\
    & $\mid$ & !e;e! & \grammarcomment{Sequential composition}\\
    & $\mid$ & !map!$_{\Gamma,x.e}$ $\mid$ !map2!$_{\Gamma,x,y.e}$ & \grammarcomment{Map and map2}\\
    & $\mid$ & !foldl!$_{\Gamma,x,y.e}$ & \grammarcomment{Fold left}\\
    \hline
    \end{tabular}
    \vspace{-0.2cm}
    \caption{Grammar of the source UNF}
    \label{fig:unf_source_grammar}
    \end{figure}

@@

 TODO: I probably need to add permutations $\sigma$ in the language of UNF then.

@@

TODO: another possible way to look at it is that UNF does some sort of localization, turning every operator to a reversible one, which is simply done by making it also return its context. I wonder if it could simply be seen as the localization of $!$. Assuming $1\times A=A$, we have $\Delta;id\times !=id$, so as $!$ would become invertible, so would be $\Delta$. This is sort of what I need in practice for reverse mode: not seeing any $\Delta$, only unary maps, and an iso can always be seen as such in this context, i.e. by making $\Delta$ an iso, it becomes essentially invisible to reverse mode. One obvious problem is that I don't know if I want all types to be isomorphic to 1.

### 2021-07-06 rad-icfp-2021/relwork.tex
minor

\noindent \textbf{Recursion}

\noindent \textbf{Higher derivatives}

\noindent \textbf{Sparse data and matrix representation}

@@

\MH{understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.}

\MH{should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}}

@@

Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
combining array languages and automatic differentiation (tensor calculus) in the best 
fitted intermediate representation for optimizations is of key interest and active research.
\cite{bernstein2020differentiating} 
\cite{laue2018computing,laue2020simple}

@@

\cite{ee2020correctness}: non diff function
\cite{sherman2021}: non diff HO functions
\cite{pearlmutter2008reverse}: based on this
\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
\cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.

>>>>

\noindent \textbf{Recursion}
TODO

\noindent \textbf{Higher derivatives}
TODO

\noindent \textbf{Sparse data and matrix representation}
TODO

@@

TODO: understand related work sentence from Diff. Curry paper: The idea of using closures as back-propagators is receiving recent attention. 
    For example Julia Zygote \cite{innes2019zygote} and Swift AD \cite{wei2018first} chose this design. 
    Other recent work follows similar ideas [27, 26] but is using meta-programming as an implementation technique.

TODO: should talk a bit about the recent paper: reverse AD in Dex \cite{paszke2021getting}

@@

Given the enormous computation needs for state-of-the-art large scale machine learning applications, 
which require extremely efficient tensor computations and automatic differentiation for backpropagation, 
combining array languages and automatic differentiation (tensor calculus) in the best 
fitted intermediate representation for optimizations is of key interest and active research.
\cite{bernstein2020differentiating} 
TODO
\cite{laue2018computing,laue2020simple}

@@

TODO
\cite{ee2020correctness}: non diff function
\cite{sherman2021}: non diff HO functions
\cite{pearlmutter2008reverse}: based on this
\cite{elliott2018simple}: bit similar because using CT as useful language, but we take impl and opti more seriously.
\cite{shaikhha2019efficient}: similar approach but here more complicated for reverse mode, would be interesting to investigate more loop fusions here.
